---
title: 'Case study'
teaching: 10
exercises: 2
---

```{r include=FALSE}
here::i_am("episodes/worked-example.Rmd")
```


:::::::::::::::::::::::::::::::::::::: questions 

- How do you write a lesson using R Markdown and `{sandpaper}`?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Explain how to use markdown with the new lesson template
- Demonstrate how to include pieces of code, figures, and nested challenge blocks

::::::::::::::::::::::::::::::::::::::::::::::::

## Peptide count data

To round things out, we'll explore some simulated peptide library data using what we've learned so far.  Please follow along in an R markdown notebook on your own computer.  The data is included in the data files you downloaded at the start of the course.

In this dataset, we've made an imaginary peptide insertion in the *cap* gene of AAV to create a plasmid library. We've then packaged this to create a vector library, and done a selection in some imaginary cells.  We've therefore got data from **plasmid**, **vector**, **DNA/entry** and **cDNA/expression** libraries.  

Imagine you've already done the counting using [the counting pipeline](https://github.com/szsctt/ngs_barcodes), and now you want to explore the results.  

You'll start with four files, one for each library, each containing two columns: *peptide* and *count*. These files are called `counts_plasmid.tsv`, `counts_vector.tsv`, `counts_entry.tsv`, `counts_expression.tsv`.

The data here is **simulated**, so don't interpret the results!  If you want to see how I simulated the data, you can check it out [on github](https://github.com/szsctt/cmri_R_workshop/blob/main/episodes/files/sim_pept.R).

## Excercies

:::::::::::::::::::::::: challenge 

Load the data into variable called `count`.  Then, create a new column called `lib` with the library type ('plasmid', 'vector', 'entry', 'expression').

:::::::::::::: solution
```{r echo=FALSE}
# load tidyverse library
library(tidyverse)
```



```{r}
# file paths
files <- here::here("episodes", "data", c("counts_plasmid.tsv", 
                                          "counts_entry.tsv",
                                          "counts_expression.tsv",
                                          "counts_vector.tsv"))

# load data
counts <- read_tsv(files, col_types = "ci", id="file")

# create extra column
counts <- counts %>% 
  mutate(lib = stringr::str_extract(file, "plasmid|vector|entry|expression")) %>% 
  select(-file)

glimpse(counts)

```

:::::::::::::::::::::::

::::::::::::::::::::::::::::::::::



:::::::::::::::::::::::: challenge 

Make a table with the following information

 - The number of distinct peptides per library
 - The mean count in each library
 - The median count in each library
 - The standard deviation of counts for each libary
 - The sum of all the counts for each library
 
:::::::::::::: solution

```{r}
counts %>% 
  group_by(lib) %>% 
  summarise(n_pept = n_distinct(peptide),
            mean_count = mean(count),
            median_count = median(count),
            sd_count = sd(count),
            lib_size = sum(count))

```

:::::::::::::::::::::::

::::::::::::::::::::::::::::::::::




:::::::::::::::::::::::: challenge

Get a table of the top 10 peptides from each library.

:::::::::::::: solution

```{r}
counts %>% 
  group_by(lib) %>% 
  slice_max(order_by=count, n=10)

```

:::::::::::::::::::::::

::::::::::::::::::::::::::::::::::




:::::::::::::::::::::::: challenge 

For each peptide, calculate the fraction of counts within the library in a new column called `frac`.

:::::::::::::: solution

```{r}
counts <- counts %>% 
  group_by(lib) %>% 
  mutate(frac = count/sum(count))

glimpse(counts)
```

:::::::::::::::::::::::

::::::::::::::::::::::::::::::::::
 




:::::::::::::::::::::::: challenge 

Reproduce this 'bubble plot' of the normalized counts.

```{r, echo=FALSE}
counts %>% 
  ggplot(aes(x=lib, y=frac, size=frac, color=lib)) +
  geom_jitter(alpha=0.5) +
  scale_y_log10() +
  labs(x="Library", y="Fraction")
```

:::::::::::::: solution

```{r}
counts %>% 
  ggplot(aes(x=lib, y=frac, size=frac, color=lib)) +
  geom_jitter(alpha=0.5) +
  scale_y_log10() +
  labs(x="Library", y="Fraction")
```

:::::::::::::::::::::::

::::::::::::::::::::::::::::::::::




:::::::::::::::::::::::: challenge 

Add column called `rank` with ranks within each library, with 1 for peptide with highest fraction, 2 for peptide with second highest, and so on.

:::::::::::::: solution
```{r}
counts <- counts %>% 
  arrange(desc(frac)) %>% 
  group_by(lib) %>% 
  mutate(rank = row_number())

glimpse(count)
```
:::::::::::::::::::::::
::::::::::::::::::::::::::::::::::



:::::::::::::::::::::::: challenge 

Make a table where there is one row for each peptide, and a column for the rank of the peptide in each library. 

That is, there's one column containing the counts for the plasmid library, one column for the counts in the vector library, and so on.

In this table, assign a unique ID (using `row_number()`) to each peptide, ranking by count in the expression library.  That is, the peptide ranked 1 in the expression library should have ID 1, the peptide ranked 2 in the expression library should have ID 2, and so on.

Assign this new table to a variable `ranked`.


:::::::::::::: solution
```{r}
ranked <- counts %>% 
  select(lib, peptide, rank) %>% 
  pivot_wider(id_cols = "peptide", names_from="lib", values_from = "rank") %>% 
  arrange(expression) %>% 
  mutate(id = row_number())

ranked
```
:::::::::::::::::::::::
::::::::::::::::::::::::::::::::::



:::::::::::::::::::::::: challenge 

Reproduce this raster plot showing the ranks for each peptide in each library
 

```{r, echo=FALSE}
ranked %>% 
  pivot_longer(expression:plasmid, names_to="lib", values_to="rank") %>% 
  ggplot(aes(x=lib, y=id, fill=rank)) +
  geom_raster() +
  scale_fill_viridis_c() +
  labs(x="Library", y="Peptide ID")
```


:::::::::::::: solution
```{r}
ranked %>% 
  pivot_longer(expression:plasmid, names_to="lib", values_to="rank") %>% 
  ggplot(aes(x=lib, y=id, fill=rank)) +
  geom_raster() +
  scale_fill_viridis_c()+
  labs(x="Library", y="Peptide ID")
```
:::::::::::::::::::::::
::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::: challenge 

Create a new table (called `expr_enrich`) in which you calculate the 'enrichment score' of each peptide in the expression library relative to the vector library (that is, $\frac{\text{fraction in expression library}}{\text{fraction in vector library}}$). Add a column called `rank` which ranks the peptides based on this enrichment score

:::::::::::::: solution

```{r}
expr_enrich <- counts %>% 
  filter(lib %in% c("expression", "vector")) %>% 
  select(peptide, lib, frac) %>% 
  pivot_wider(names_from = lib, values_from = frac) %>% 
  mutate(enrichment = expression / vector) %>% 
  arrange(desc(enrichment)) %>% 
  mutate(rank = row_number())
 

expr_enrich 
```


:::::::::::::::::::::::
::::::::::::::::::::::::::::::::::



:::::::::::::::::::::::: challenge 

Do the same, but comparing the vector to plasmid library to get an idea of how well each peptide packages. Filter for rows where the 'enrichment score' is less than 1 to get a list of 'poor packagers'.  
Save this to a variable called `vec_enrich`.

:::::::::::::: solution

```{r}
vec_enrich <- counts %>% 
  filter(lib %in% c("plasmid", "vector")) %>% 
  select(peptide, lib, frac) %>% 
  pivot_wider(names_from = lib, values_from = frac) %>% 
  mutate(enrichment = vector / plasmid) %>% 
  arrange(desc(enrichment)) %>% 
  mutate(rank = row_number())

poor_packagers <- vec_enrich %>% 
  filter(enrichment < 1)

poor_packagers
```


:::::::::::::::::::::::
::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::: challenge 

Make a table of the top 500 peptides ranked by expression enrichment, removing any of the poor packagers.  Save this table to a file called `good_peptides.tsv`.

:::::::::::::: solution

```{r}
good_peptides <- expr_enrich %>% 
  filter(!peptide %in% poor_packagers$peptide) %>% 
  slice_max(order_by=enrichemnt, n=500)

good_peptides 

write_tsv(good_peptides, file=here::here("good_peptides.tsv"))
```


:::::::::::::::::::::::
::::::::::::::::::::::::::::::::::



::::::::::::::::::::::::::::::::::::: keypoints 

- Use `.md` files for episodes when you want static content
- Use `.Rmd` files for episodes when you need to generate output
- Run `sandpaper::check_lesson()` to identify any issues with your lesson
- Run `sandpaper::build_lesson()` to preview your lesson locally

::::::::::::::::::::::::::::::::::::::::::::::::

